---
title: "Interactive Session 2B"
subtitle: "Higher order derivatives, partial derivatives, plotting functions in ggplot"
editor_options: 
  chunk_output_type: console
---

## 1. Partial derivatives by hand

a. Find all partial derivatives of the following: 

$$f(a, z, t)=3at^2-4t+5.1az-z^3$$

::: {.callout-tip collapse=true}
## Solution
$\frac{\partial f}{\partial a} = 3t^2 + 5.1z$

$\frac{\partial f}{\partial z} = 5.1a - 3z^2$

$\frac{\partial f}{\partial t} = 6at - 4$
:::

b. Algal biomass ($B$, grams) in a system is modeled by the following, where $h$ is time (hours) and $T$ is temperature (Celsius). 

$$B(h, T)=0.6Th+2.1h^2$$
For water at 28 degrees, find an expression for how biomass is changing with respect to time ($\frac{\partial B}{\partial h}$). 

::: {.callout-tip collapse=true}
## Solution
$\frac{\partial B}{\partial h}=0.6T + (2)(2.1)h$

$\frac{\partial B}{\partial h}=(0.6)(28) + 4.2h$

$\frac{\partial B}{\partial h}=16.8 + 4.2h$
:::

c. Find the 3^rd^ derivative of the following: 

$$G(t)=2.2 + 3.1t-5.6t^4 + \frac{4.3}{t^3}$$

::: {.callout-tip collapse=true}
## Solution
**First derivative:**

  - $G'(t) = 3.1 -(4)(5.6)t^3 +(-3)(4.3)t^{-4}$
  
  - $G'(t) = 3.1 -22.4t^3 -12.9t^{-4}$

**Second derivative:**

  - $G''(t) = 3.1. - (22.4)(3)t^2 -(12.9)(-4)t^{-5}$
  
  - $G''(t) = 3.1. - 67.2t^2 -(12.9)(-4)t^{-5}$
  
  - $G''(t) = -67.2t^2 + 51.6t^{-5}$
  
**Third derivative:**

  - $G'''(t) = -(67.2)(2)t + (51.6)(-5)t^{-6}$
  
  - $G'''(t) = -134.4t - 258t^{-6}$
:::

## 2. Plotting a function using `{ggplot2}`

- Add a new Quarto doc (I'm saving mine as `day2b-exercises.qmd`) to your existing `eds212-day2-GHpractice` repository and follow along with the exercises, below. 

#### Example 1:

::: {.callout-tip collapse=true}
## Use keyboard shortcuts to create a function in R
In the code chunk below, we create an R function to plot the function, $f(x) = 3x^2 + 4$. We can use a nifty keyboard shortcut to create a R function around this expression following these steps:

1. Write out the expression $3x^2 + 4$ in a code chunk:

```{r}
#| eval: false
#| echo: true
3 * x^2 + 4
```

2. Highlight the expression, then use the keyboard shortcut `control` + `option` + `X` (Mac) or `Ctrl` + `Alt` + `X` (Windows) to pop open a dialog box, which will prompt you to provide a name for your function (you can call it `eq1`).

3. Watch your function appear! The shortcut will automatically add any variables from the body of your function (here, just `x`) as arguments inside `function()`:

```{r}
#| eval: false
#| echo: true
eq1 <- function(x) {
  3 * x^2 + 4
}
```

:::

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
# load packages ----
library(tidyverse)

# Define function ----
eq1 <- function(x) {
  3 * x^2 + 4
} 

# create data frame with range of values to evaluate function over ----
my_data_range <- data.frame(x = c(1, 100))

# Plot it as a continuous curve over a specific range using `geom_function()` ----
# `geom_function()` automatically evaluates the function over the range of x-values specified in the df and adds the resulting y-values to the plot
ggplot(data = my_data_range, aes(x = x)) + 
  geom_function(fun = eq1)
```

#### Example 2:

```{r}
#| eval: true
#| echo: true
# Define function ----
eq2 <- function(x) {
  2.4 - 5*x^3 + 2.1*x^2
} 

# Plot it as a continuous curve over a specific range using `geom_function()` (here, we define our data frame directly inside `ggplot()` ----
ggplot(data = data.frame(x = c(-50, 50)), aes(x = x)) + 
  geom_function(fun = eq2)
```

## 3. Plotting derivatives

Let's look at the function: $C(t)=t^3$

Let's plot the function along with the first derivative: 

```{r}
#| eval: true
#| echo: true
# Store the function C(t) ----
ct <- function(t) {
  t^3
}

# create data frame with range of values to evaluate function over ----
my_data_range <- data.frame(t = c(-5, 5))

# Plot it ----
ggplot(data = my_data_range, aes(x = t)) + 
  geom_function(fun = ct)

# Find the derivative ----     
dc_dt <- D(expr = expression(t^3), name = "t")
dc_dt

# Store the derivative as a function ----
dc_dt_fun <- function(t) {
  3 * t ^2
}

# Then plot them together ----
ggplot(data = my_data_range, aes(x = t)) + 
  geom_function(fun = ct, color = "red") +
  geom_function(fun = dc_dt_fun, color = "blue")
```

## 4. Partial derivatives in R

Specify the variable you want to take the derivative with respect to. 

$$f(x,y,z)=2xy-3x^2z^3$$

Find all partials, $\frac{\partial f}{\partial x}$, $\frac{\partial f}{\partial y}$ and $\frac{\partial f}{\partial z}$

```{r}
#| eval: true
#| echo: true
# Create the expression ----
f_xyz <- expression(2 * x * y - 3 * x^2 * z^3)

# Partial with respect to x ----
df_dx <- D(expr = f_xyz, name = "x")
df_dx

# Partial with respect to y ----
df_dy <- D(expr = f_xyz, name = "y")
df_dy

# Partial with respect to z ---- 
df_dz <- D(expr = f_xyz, name = "z")
df_dz
```

## 5. One more example:

Given the function: $$P(q)=2cos(3q+0.4)-5.6e^{1.4q}$$ 

Find the instantaneous slope at $q = 0.8$.

```{r}
# First, create the expression ---- 
pq <- expression(2 * cos(3 * q + 0.4) - 5.6 * exp(1.4 * q))

# Find the first derivative with respect to q ----
dp_dq <- D(expr = pq, name = "q")

# Return the first derivative ----
dp_dq

# Define value of q ----
q <- 0.8

# Evaluate dp_dq at that value ----
eval(dp_dq)
```

## 6. Stage, commit, (pull) and push your changes to GitHub

Follow the same pattern that we practiced in today's earlier [interactive session (2A)](interactive-session-2a.qmd#hello-git-github), and verify that it worked by checking out your GitHub repo.  


<br>

::: {.center-text .body-text-xl .teal-text}
End interactive session 2B
:::