---
format: 
  revealjs:
    slide-number: true
    # code-link: true
    highlight-style: a11y
    chalkboard: true
    theme: 
      - ../../meds-slides-styles.scss
---

## {#title-slide data-menu-title="Title Slide" background="#053660"} 

[EDS 212: Day 3, Lecture 2]{.custom-title}

[*Linear algebra fundamentals*]{.custom-subtitle}

<hr class="hr-teal">

[August 7^th^, 2024]{.custom-subtitle3}

---

## {#intro data-menu-title="Intro"} 

[Linear algebra introduction]{.slide-title}

<hr>

<br>

::: {.body-text-m}
Linear algebra "branch of mathematics concerning linear equations" (Wikipedia), sometimes also described as the math of vectors & matrices. 

<br>

It is a fundamental part of data science (and how computers understand & process data), and useful for describing environmental processes.

<br>

As [Dr. Jason Brownlee](https://machinelearningmastery.com/gentle-introduction-linear-algebra/) writes, "Linear algebra is the mathematics of data."
:::

---

## {#building-blocks data-menu-title="Building blocks"} 

[The building blocks of linear algebra]{.slide-title}

<hr>

<br>

::: {.body-text-l}
- **scalar:** a value without direction, representing magnitude

- **vector:** an ordered list of values, representing magnitude *and* direction (physics) or values for an observation or variable (data science)

- **matrix:** a table of values
:::

---

## {#applications data-menu-title="Applications"} 

[Applications of linear algebra in environmental sciences]{.slide-title2}

<hr>

<br>

::: {.body-text-m}
- Dimensional reduction
- Population matrix models
- Optimization

<br>

**Applications of linear algebra in data science:**

- Array programming / vectorized code
- Machine learning
- For loops
:::

---

## {#vectors data-menu-title="Vectors"} 

[Let's start with vectors]{.slide-title}

<hr>

<br>

[**Where are vectors in EDS?**]{.body-text-l}

. . . 

<br>

[***Everywhere.*** ]{.body-text-l}

<br>

::: {.body-text-m}
Vectors are lists of values used to describe different features or variables of interest. For example, if you are trying to model *fish size* based on length (cm) and mass (g), then for a fish with length 32 cm weighing 281 g, you might describe that by: 

$$\vec{F} = [32, 281]$$
:::

---

## {#df data-menu-title="data frames"} 

[From Wickham's *Advanced R* chapter on [Data Structures](http://adv-r.had.co.nz/Data-structures.html):]{.slide-title2}

<hr>

<br>

::: {.body-text-l}
>"Under the hood, a data frame is a list of equal-length vectors."
:::

```{r}
#| eval: true
#| echo: true
food <- c("banana", "apple", "carrot")
meal <- c("breakfast", "snack", "lunch")
food_mass_g <- c(14.8, 19.2, 11.5)

squirrel_meals <- data.frame(food, meal, food_mass_g) 
squirrel_meals # Returns the data frame
```

---

## {#notation data-menu-title="Notation"} 

[Notation]{.slide-title}

<hr>

<br>

::: {.body-text-l}
Vectors are indicated with an arrow over the vector name:

<br>

$\vec{A} = [1,5]$

<br>

$\vec{AB} = [10.2, 3.1]$
:::

---

## {#add-sub data-menu-title="Addition & subtraction"} 

[Vector addition & subtraction]{.slide-title}

<hr>

::: {.body-text-l}
Just add or subtract the corresponding pieces.

<br>

If: $\vec{A} = [2, 6]$ and $\vec{B} = [11, 10]$, then: 

<br>

$$\vec{A} - \vec{B} = [2-11, 6-10] = [-9, -4]$$

<br>

What does this look like graphically? **Let's draw it!**
:::

---

## {#scalar-multipliers data-menu-title="Scalar multipliers"} 

[Scalar multipliers]{.slide-title}

<hr>

<br>

::: {.body-text-m}
You can multiply any vector by a scalar (constant). This will not change the *direction* of the vector - it will only change the *magnitude* of the vector. 

<br>

**Example:** $\vec{A} = [2,4]$

<br>

$$\vec{B} = 3\vec{A} = [3*2, 3*4] = [6, 12]$$

<br>

What does this look like graphically? **Let's draw it!**
:::

---

## {#draw-in-r data-menu-title="Draw in R"} 

[In R:]{.slide-title}

<hr>

<br>

[We create vectors in R using the `c()` function (for combine or concatenate), and can perform operations on numeric vectors using basic operators.]{.body-text-l}

<br>

```{r}
#| eval: true
#| echo: true
# Create vectors A and B ----
A <- c(1,2)
B <- c(5,9)
```

---

## {#draw-in-r data-menu-title="Add"} 

[Vector addition and subtraction in R]{.slide-title}

<hr>

<br>

::: {.body-text-l}
Just do it! 

<br>

For $\vec{A}=[1,2]$ and $\vec{B}=[5,9]$:
:::

<br>

```{r}
#| eval: true
#| echo: true
# Addition ----
A + B

# Subtraction ----
A - B
```

---

## {#scalar-multiplication data-menu-title="Scalar multiplication"} 

[Scalar multiplication in R]{.slide-title}

<hr>

<br>

::: {.body-text-l}
Just do it!

<br>

For $\vec{A}=[1,2]$, calculate $4\vec{A}$:
:::

```{r}
#| eval: true
#| echo: true
4 * A
```

---

## {#vector-3 data-menu-title="Vectors > 3 elements"} 

[Vectors of > 3 elements]{.slide-title}

<hr>

<br>

::: {.body-text-l}
$\vec{M}=c(2,4,1,8,6)$

<br>

Is as valid as describing a "point" in multivariate space as a vector with two "coordinates" -- it's just difficult for us to visualize and conceptualize since our brain only happily deals with 3 dimensions. 
:::

---

## {#dot-product data-menu-title="Dot product"} 

[Dot product]{.slide-title}

<hr>

<br>

::: {.body-text-l}
For vectors $\vec a$ and $\vec b$, their dot product is:

<br>

$$\vec a \cdot \vec b = \sum a_i b_i$$
:::

<br>

[[**In words:**]{.body-text-m} The dot product is the sum of elements of each vector multiplied together, and is a measure of how close the vectors "point" in the same direction]{.body-text-m}

---

## {#dot-product-example data-menu-title="Dot product example"} 

[Dot product example]{.slide-title}

<hr>

<br>

::: {.body-text-m}
For $\vec a=[2,-1,0]$ and $\vec b=[9,3,-4]$:

<br>

$$\vec a \cdot \vec b = (2)(9)+(-1)(3)+(0)(-4) =15$$

<br>

This becomes very useful when describing systems of equations (tomorrow).
:::

---

## {#orthogonal-vectors data-menu-title="Orthogonal vectors"} 

[What happens when we have orthogonal vectors?]{.slide-title2}

<hr>

<br>

::: {.body-text-m}
Sketch a quick graph, then find the dot product, of the following vector combinations: 

<br>

1. $\vec a=[0,4]$ and $\vec b =[6,0]$

<br>

2. $\vec c=[-3,1]$ and $\vec d=[2,6]$

<br>

What is the value of the dot product for orthogonal vectors? 
:::

---

## {#vec-fundamentals data-menu-title="Vector fundamentals"} 

[More on vector fundamentals:]{.slide-title}

<hr>

<br>
<br>
<br>
<br>

[**Optional:** watch 3Brown1Blue's great 10 min recording on [Vectors (Ch 1 Essense of Linear Algebra)](https://www.youtube.com/watch?v=fNk_zzaMoSs&list=PLZHQObOWTQDPD3MizzM2xVFitgF8hE_ab&index=1).]{.body-text-m}

<br>

::: {.center-text .body-text-l .teal-text}
Vector addition & scalar multiplication are the basis of most linear algebra! 
:::


